\documentclass[main.tex]{subfiles}
\begin{document}

\section{Introduction}
This project seeks to provide means for translating from a subset of English
into Overpass queries using a Combinatory Categorial Grammar.

\subsection{\fontfamily{comic}\selectfont Motivation}
The subject of this thesis came into being rather randomly.
First, the following ingredients were implanted into a cadaver:
\begin{itemize}
    \item My affection towards maps and locations
    \item My contempt towards \sout{black boxes invented by aliens}
        machine learning
    \item My desire to do something with $\lambda$-calculus / functional
        languages / compiling
    \item My having read the wiki page about Overpass several days before
        thinking about a thesis subject
\end{itemize}
Then, said cadaver was strapped on a table in a\footnote{The tower of
The Faculty of Chemistry and Pharmacy of St. Kliment Ohridski University
of Sofia, located at 42.67455,23.33225.} tower and its limbs connected
to the tower's lightning rod. The lightning that figuratively struck the
lightning rod came as a \emph{sample subject} presented by my (as of now)
supervisor for this thesis --- something along the lines of ``Translating
a subset of a natural language into a formal language (SQL or Haskell)
via Combinatory Categorial Grammar''. After refracting the aforementioned
ingredients through the plasmatic halo of this sample subject, this thesis
was implemented, with no regard for any usefulness or state-of-the-art--ness.
The assessment of such qualities is left as exercise to the reader.

\subsection{Motivation}
It is often the case that humans need to communicate a concrete question
or order (a query) to a computer. The most straightforward way for this to happen is
to have the human write out their query in a domain-specific language (DSL),
specially crafted to handle the given type of query. This approach, albeit
very efficient when done correctly, requires substantial upfront effort:
the human needs to get familiar with the DSL and must have access to a specific
intermediary
(usually a keyboard and a screen). Thus, in some cases it is desirable
to form an expression in a natural language (or a language very similar to a
subset of a natural language) and have it translated to a query in the given
DSL.

The point is not to translate \emph{any} sentence in the source language
that has semantics applicable to the target DSL, but rather to define a
subset of the source language, in which a human could express a query
reasonably easily.

Most DSLs have compositional semantics, which makes them great candidates
for generation using Combinatory Category Grammars (CCGs) \cite[p.~181]{nts}.

One such DSL is the Overpass language, used by the open source project
OpenStreetMap for making queries to the map database. It has been chosen as
a concrete target language for this project in order to assess the usefulness
of CCG for translating from a natural language subset to a compositional
DSL.
\end{document}
